A Bioconductor Workflow for Processing and Analysing Spatial Proteomics Data

### 2016-12-03 bioc-workflow.tex
add Jean Beltran reference

Quantitative mass spectrometry based spatial proteomics involves
elaborate, expensive and time consuming experimental protocols and
considerable effort is invested in the generation of such data.
Multiple research groups have described a variety of approaches to
establish high quality proteome-wide datasets (see for example
\cite{Gatto:2010} for a review, and \cite{hyper,Itzhak:2016} for
recent examples). However, data analysis is as critical as data
production for reliable and insightful biological
interpretation. Here, we walk the reader through a typical pipeline
for the analysis of such data using several Bioconductor packages for
the R statistical programming environment.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{combined} \hlkwb{<-} \hlkwd{combine}\hlstd{(hyperLOPIT2015ms3r1, hyperLOPIT2015ms3r2)}
\hlstd{combined}
\end{alltt}
\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 6725 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126.1 X127N.1 ... X131.2 (20 total)
##   varLabels: Replicate TMT.Reagent ... Iodixonal.Density (5 total)
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2Y3-3 (6725 total)
##   fvarLabels: EntryName ProteinDescription ... ProteinCoverage2 (8
##     total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Combined [6725,20] and [6268,10] MSnSets Sat Dec  3 19:46:20 2016 
##  MSnbase version: 1.21.7
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{combined} \hlkwb{<-} \hlkwd{filterNA}\hlstd{(combined)}
\hlstd{combined}
\end{alltt}
\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 5032 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126.1 X127N.1 ... X131.2 (20 total)
##   varLabels: Replicate TMT.Reagent ... Iodixonal.Density (5 total)
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2R6 (5032 total)
##   fvarLabels: EntryName ProteinDescription ... ProteinCoverage2 (8
##     total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Combined [6725,20] and [6268,10] MSnSets Sat Dec  3 19:46:20 2016 
## Subset [6725,20][5032,20] Sat Dec  3 19:46:21 2016 
## Removed features with more than 0 NAs: Sat Dec  3 19:46:21 2016 
## Dropped featureData's levels Sat Dec  3 19:46:21 2016 
##  MSnbase version: 1.21.7
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{figure}[!ht]
  \centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{prots} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"Q8CG48"}\hlstd{,} \hlstr{"Q8CG47"}\hlstd{,} \hlstr{"Q8K2Z4"}\hlstd{,} \hlstr{"Q8C156"}\hlstd{)}
\hlstd{foi13s} \hlkwb{<-} \hlkwd{FeaturesOfInterest}\hlstd{(}\hlkwc{description} \hlstd{=} \hlstr{"13S consensin proteins"}\hlstd{,}
                             \hlkwc{fnames} \hlstd{= prots,}
                             \hlkwc{object} \hlstd{= hl)}
\hlstd{foi13s}
\end{alltt}
\begin{verbatim}
## Traceable object of class "FeaturesOfInterest"
##  Created on Sat Dec  3 19:46:22 2016 
##  Description:
##   13S consensin proteins
##  4 features of interest:
##    Q8CG48, Q8CG47, Q8K2Z4, Q8C156
\end{verbatim}
\begin{alltt}
\hlkwd{plot2D}\hlstd{(hl)}
\hlkwd{addLegend}\hlstd{(hl,} \hlkwc{cex} \hlstd{=} \hlnum{.6}\hlstd{)}
\hlkwd{highlightOnPlot}\hlstd{(hl, foi13s)}
\end{alltt}
\end{kframe}
\includegraphics[width=.8\textwidth]{figure/foi-1}

@@

% latex table generated in R 3.3.2 by xtable 1.8-2 package
% Sat Dec  3 19:46:23 2016
\begin{table}[ht]
\centering
\begin{tabular}{rrllr}
  \hline
 & Replicate & Tag & Gradient.Fraction & Iodixonal.Density \\ 
  \hline
X127C &   1 & 127C & 8 to 9 (pooled) & 11.00 \\ 
  X127C.1 &   2 & 127C & 7 to 9 (pooled) & 10.00 \\ 
   \hline
\end{tabular}
\caption{Differences in gradient fraction pooling.} 
\label{tab:pdtab}
\end{table}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{hl}
\end{alltt}
\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 5032 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126 X127N ... X131.1 (20 total)
##   varLabels: Replicate Tag Gradient.Fraction Iodixonal.Density
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2R6 (5032 total)
##   fvarLabels: uniprot.accession uniprot.id ... pd (27 total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Added markers from  'mrk' marker vector. Sat Dec  3 19:46:22 2016 
## Added markers from  'pdres' marker vector. Sat Dec  3 19:46:23 2016 
##  MSnbase version: 2.1.4
\end{verbatim}
\begin{alltt}
\hlkwd{getMarkers}\hlstd{(hl,} \hlkwc{fcol} \hlstd{=} \hlstr{"pd"}\hlstd{)}
\end{alltt}
\begin{verbatim}
## organelleMarkers
##                          40S Ribosome 
##                                   106 
##                          60S Ribosome 
##                                    95 
## Endoplasmic reticulum/Golgi apparatus 
##                                   393 
##                         Mitochondrion 
##                                   525 
##                           Phenotype 1 
##                                   300 
##                           Phenotype 2 
##                                   253 
##                           Phenotype 3 
##                                   203 
##                           Phenotype 4 
##                                    74 
##                           Phenotype 5 
##                                    91 
##                       Plasma membrane 
##                                   421 
##                            Proteasome 
##                                    92 
##                               unknown 
##                                  2479
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{(ts} \hlkwb{<-} \hlkwd{orgQuants}\hlstd{(hl,} \hlkwc{fcol} \hlstd{=} \hlstr{"svm"}\hlstd{,} \hlkwc{scol} \hlstd{=} \hlstr{"svm.scores"}\hlstd{,} \hlkwc{mcol} \hlstd{=} \hlstr{"markers"}\hlstd{,} \hlkwc{t} \hlstd{=} \hlnum{.5}\hlstd{))}
\end{alltt}
\begin{verbatim}
##            40S Ribosome            60S Ribosome      Actin cytoskeleton 
##               0.4332618               0.3058476               0.3871539 
##                 Cytosol   Endoplasmic reticulum                Endosome 
##               0.6745670               0.6093870               0.4335865 
##    Extracellular matrix                Lysosome           Mitochondrion 
##               0.4193893               0.6121402               0.9498369 
##     Nucleus - Chromatin Nucleus - Non-chromatin              Peroxisome 
##               0.7946622               0.7092443               0.3158764 
##         Plasma membrane              Proteasome 
##               0.7096980               0.4169878
##            40S Ribosome            60S Ribosome      Actin cytoskeleton 
##               0.4332618               0.3058476               0.3871539 
##                 Cytosol   Endoplasmic reticulum                Endosome 
##               0.6745670               0.6093870               0.4335865 
##    Extracellular matrix                Lysosome           Mitochondrion 
##               0.4193893               0.6121402               0.9498369 
##     Nucleus - Chromatin Nucleus - Non-chromatin              Peroxisome 
##               0.7946622               0.7092443               0.3158764 
##         Plasma membrane              Proteasome 
##               0.7096980               0.4169878
\end{verbatim}
\begin{alltt}
\hlstd{hl} \hlkwb{<-} \hlkwd{getPredictions}\hlstd{(hl,} \hlkwc{fcol} \hlstd{=} \hlstr{"svm"}\hlstd{,} \hlkwc{scol} \hlstd{=} \hlstr{"svm.scores"}\hlstd{,} \hlkwc{mcol} \hlstd{=} \hlstr{"markers"}\hlstd{,} \hlkwc{t} \hlstd{= ts)}
\end{alltt}
\begin{verbatim}
## ans
##            40S Ribosome            60S Ribosome      Actin cytoskeleton 
##                      85                     172                      85 
##                 Cytosol   Endoplasmic reticulum                Endosome 
##                     298                     475                     107 
##    Extracellular matrix                Lysosome           Mitochondrion 
##                      25                     119                     523 
##     Nucleus - Chromatin Nucleus - Non-chromatin              Peroxisome 
##                     230                     342                      38 
##         Plasma membrane              Proteasome                 unknown 
##                     320                     157                    2056
\end{verbatim}
\end{kframe}
\end{knitrout}

>>>>

Quantitative mass spectrometry based spatial proteomics involves
elaborate, expensive and time consuming experimental protocols and
considerable effort is invested in the generation of such data.
Multiple research groups have described a variety of approaches to
establish high quality proteome-wide datasets (see for example
\cite{Gatto:2010} for a review, and
\cite{hyper,Itzhak:2016,Jean_Beltran:2016} for recent
examples). However, data analysis is as critical as data production
for reliable and insightful biological interpretation. Here, we walk
the reader through a typical pipeline for the analysis of such data
using several Bioconductor packages for the R statistical programming
environment.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{combined} \hlkwb{<-} \hlkwd{combine}\hlstd{(hyperLOPIT2015ms3r1, hyperLOPIT2015ms3r2)}
\hlstd{combined}
\end{alltt}
\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 6725 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126.1 X127N.1 ... X131.2 (20 total)
##   varLabels: Replicate TMT.Reagent ... Iodixonal.Density (5 total)
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2Y3-3 (6725 total)
##   fvarLabels: EntryName ProteinDescription ... ProteinCoverage2 (8
##     total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Combined [6725,20] and [6268,10] MSnSets Sat Dec  3 20:38:06 2016 
##  MSnbase version: 1.21.7
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{combined} \hlkwb{<-} \hlkwd{filterNA}\hlstd{(combined)}
\hlstd{combined}
\end{alltt}
\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 5032 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126.1 X127N.1 ... X131.2 (20 total)
##   varLabels: Replicate TMT.Reagent ... Iodixonal.Density (5 total)
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2R6 (5032 total)
##   fvarLabels: EntryName ProteinDescription ... ProteinCoverage2 (8
##     total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Combined [6725,20] and [6268,10] MSnSets Sat Dec  3 20:38:06 2016 
## Subset [6725,20][5032,20] Sat Dec  3 20:38:06 2016 
## Removed features with more than 0 NAs: Sat Dec  3 20:38:06 2016 
## Dropped featureData's levels Sat Dec  3 20:38:06 2016 
##  MSnbase version: 1.21.7
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{figure}[!ht]
  \centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{prots} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"Q8CG48"}\hlstd{,} \hlstr{"Q8CG47"}\hlstd{,} \hlstr{"Q8K2Z4"}\hlstd{,} \hlstr{"Q8C156"}\hlstd{)}
\hlstd{foi13s} \hlkwb{<-} \hlkwd{FeaturesOfInterest}\hlstd{(}\hlkwc{description} \hlstd{=} \hlstr{"13S consensin proteins"}\hlstd{,}
                             \hlkwc{fnames} \hlstd{= prots,}
                             \hlkwc{object} \hlstd{= hl)}
\hlstd{foi13s}
\end{alltt}
\begin{verbatim}
## Traceable object of class "FeaturesOfInterest"
##  Created on Sat Dec  3 20:38:08 2016 
##  Description:
##   13S consensin proteins
##  4 features of interest:
##    Q8CG48, Q8CG47, Q8K2Z4, Q8C156
\end{verbatim}
\begin{alltt}
\hlkwd{plot2D}\hlstd{(hl)}
\hlkwd{addLegend}\hlstd{(hl,} \hlkwc{cex} \hlstd{=} \hlnum{.6}\hlstd{)}
\hlkwd{highlightOnPlot}\hlstd{(hl, foi13s)}
\end{alltt}
\end{kframe}
\includegraphics[width=.8\textwidth]{figure/foi-1}

@@

% latex table generated in R 3.3.2 by xtable 1.8-2 package
% Sat Dec  3 20:38:08 2016
\begin{table}[ht]
\centering
\begin{tabular}{rrllr}
  \hline
 & Replicate & Tag & Gradient.Fraction & Iodixonal.Density \\ 
  \hline
X127C &   1 & 127C & 8 to 9 (pooled) & 11.00 \\ 
  X127C.1 &   2 & 127C & 7 to 9 (pooled) & 10.00 \\ 
   \hline
\end{tabular}
\caption{Differences in gradient fraction pooling.} 
\label{tab:pdtab}
\end{table}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{hl}
\end{alltt}
\begin{verbatim}
## MSnSet (storageMode: lockedEnvironment)
## assayData: 5032 features, 20 samples 
##   element names: exprs 
## protocolData: none
## phenoData
##   sampleNames: X126 X127N ... X131.1 (20 total)
##   varLabels: Replicate Tag Gradient.Fraction Iodixonal.Density
##   varMetadata: labelDescription
## featureData
##   featureNames: Q9JHU4 Q9QXS1-3 ... Q9Z2R6 (5032 total)
##   fvarLabels: uniprot.accession uniprot.id ... pd (27 total)
##   fvarMetadata: labelDescription
## experimentData: use 'experimentData(object)'
## Annotation:  
## - - - Processing information - - -
## Added markers from  'mrk' marker vector. Sat Dec  3 20:38:08 2016 
## Added markers from  'pdres' marker vector. Sat Dec  3 20:38:09 2016 
##  MSnbase version: 2.1.4
\end{verbatim}
\begin{alltt}
\hlkwd{getMarkers}\hlstd{(hl,} \hlkwc{fcol} \hlstd{=} \hlstr{"pd"}\hlstd{)}
\end{alltt}
\begin{verbatim}
## organelleMarkers
##                          40S Ribosome 
##                                   106 
##                          60S Ribosome 
##                                    95 
## Endoplasmic reticulum/Golgi apparatus 
##                                   393 
##                         Mitochondrion 
##                                   525 
##                           Phenotype 1 
##                                   300 
##                           Phenotype 2 
##                                   253 
##                           Phenotype 3 
##                                   203 
##                           Phenotype 4 
##                                    74 
##                           Phenotype 5 
##                                    91 
##                       Plasma membrane 
##                                   421 
##                            Proteasome 
##                                    92 
##                               unknown 
##                                  2479
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlstd{(ts} \hlkwb{<-} \hlkwd{orgQuants}\hlstd{(hl,} \hlkwc{fcol} \hlstd{=} \hlstr{"svm"}\hlstd{,} \hlkwc{scol} \hlstd{=} \hlstr{"svm.scores"}\hlstd{,} \hlkwc{mcol} \hlstd{=} \hlstr{"markers"}\hlstd{,} \hlkwc{t} \hlstd{=} \hlnum{.5}\hlstd{))}
\end{alltt}
\begin{verbatim}
##            40S Ribosome            60S Ribosome      Actin cytoskeleton 
##               0.4347094               0.3030356               0.3854031 
##                 Cytosol   Endoplasmic reticulum                Endosome 
##               0.6849680               0.5994838               0.4265235 
##    Extracellular matrix                Lysosome           Mitochondrion 
##               0.4183489               0.5905225               0.9493973 
##     Nucleus - Chromatin Nucleus - Non-chromatin              Peroxisome 
##               0.7954761               0.7092834               0.3116761 
##         Plasma membrane              Proteasome 
##               0.7196883               0.4190620
##            40S Ribosome            60S Ribosome      Actin cytoskeleton 
##               0.4347094               0.3030356               0.3854031 
##                 Cytosol   Endoplasmic reticulum                Endosome 
##               0.6849680               0.5994838               0.4265235 
##    Extracellular matrix                Lysosome           Mitochondrion 
##               0.4183489               0.5905225               0.9493973 
##     Nucleus - Chromatin Nucleus - Non-chromatin              Peroxisome 
##               0.7954761               0.7092834               0.3116761 
##         Plasma membrane              Proteasome 
##               0.7196883               0.4190620
\end{verbatim}
\begin{alltt}
\hlstd{hl} \hlkwb{<-} \hlkwd{getPredictions}\hlstd{(hl,} \hlkwc{fcol} \hlstd{=} \hlstr{"svm"}\hlstd{,} \hlkwc{scol} \hlstd{=} \hlstr{"svm.scores"}\hlstd{,} \hlkwc{mcol} \hlstd{=} \hlstr{"markers"}\hlstd{,} \hlkwc{t} \hlstd{= ts)}
\end{alltt}
\begin{verbatim}
## ans
##            40S Ribosome            60S Ribosome      Actin cytoskeleton 
##                      85                     170                      87 
##                 Cytosol   Endoplasmic reticulum                Endosome 
##                     300                     472                     102 
##    Extracellular matrix                Lysosome           Mitochondrion 
##                      27                     124                     523 
##     Nucleus - Chromatin Nucleus - Non-chromatin              Peroxisome 
##                     229                     345                      38 
##         Plasma membrane              Proteasome                 unknown 
##                     317                     155                    2058
\end{verbatim}
\end{kframe}
\end{knitrout}
